# Testing your code

Testing code is crucial, and we all do it in some form or another. The problem
is that it is not something that we do consistently. Usually code gets tested in
the beginning of a project, but then, as we start focusing on the analysis more
and more and need to respect deadlines, testing gets forgotten.

In this chapter, you are going to learn how to make testing your code consistent
and, very importantly, fully automatic. Just like in the previous chapter, the
key is to *write everything down*. Don’t just do a little test in the console to
see if the function you’ve just written works as expected. Write it down! And
don’t rely on future you to run tests, because future you is just as unreliable
as you are. Tests need to be run each time *any* of the code from a project gets
changed. This might seem overkill (why test a function that you didn’t even
touch for weeks?), because ther are dependencies between your functions, a
change in one function can affect another. Especially if the output of function
A is the input of function B, and now, because of your changes, the output of
function A changed in a way that it breaks function B, or also modifies its
input in an unexpected way.

There are several types of tests that we can use:

- unit testing: these are written while developing, and executed while developing;
- assertive testing: these are executed at runtime. These make sure, for example, that the inputs a function receives are sane.

## Unit testing

Unit testing is the testing of units. What’s a unit? Functions are units! We acutally
already ecountered one unit test before, in the `save_data.Rmd` script:

````{verbatim}

```{r tests-clean_flat_data}
# We now need to check if we have them all in the data. The test needs to be self-contained, hence
# why we need to redefine the required variables:

former_communes <- get_former_communes()

current_communes <- get_current_communes()

communes <- get_test_communes(former_communes, current_communes)

raw_data <- get_raw_data(url = "https://github.com/b-rodrigues/rap4all/raw/master/datasets/vente-maison-2010-2021.xlsx")

flat_data <- clean_raw_data(raw_data)

testthat::expect_true(
            all(communes %in% unique(flat_data$locality))
                      )
```

`````

When using `{fusen}`, a unit test should be a self-contained chunk that can be
executed completely independently. This is why in this chunk we re-created the
different variables that we needed, `communes` and the `flat_data`. If you were
developing the package without `{fusen}`, you would do the same, so don’t think
that this is somehow a limitation of `{fusen}`.

The test above ensures that we find all the former and current communes of
Luxembourg in our dataset. Let me explain again why we want to write such a test
down in a script and not simply try it out in our console. 

For this test to pass, a lot of moving pieces have to fall together. If anything
changes, be it because you changed something in either `get_raw_data()` or
`clean_raw_data()` or because something changed with the Wikipedia tables you
scraped, this test will not pass. And you should be made aware of failures as
soon as possible! Also, this test ensures that when the data gets updated, you
are certain that if you use the code in `save_data.Rmd` you will get a new
dataset that is likely correct, even if new communes merge. And mergers will
happen around 2024 by the way, the communes of Groussbous and Wal will merge,
and the communes of Bous and Waldbredimus as well. So you need to make sure that
when this happens, your code knows how to handle this, or at least gives out an
error as early as possible.

Ideally, we would test every function that we wrote, but sometimes that’s not
really possible, either due to lack of time, or because the function is quite
trivial, so maybe no test is warranted. But be careful what you consider trivial
though, I have personally been bitten in the past by trivially simple
functions! For example, a function like this one:

````{verbatim}
```{r function-make_commune_level_data}
#' make_commune_level_data Makes the final data at commune level
#'
#' @param flat_data Flat data df as returned by clean_flat_data()
#' @importFrom dplyr filter
#' @return A data frame
#' @export
make_commune_level_data <- function(flat_data){
  flat_data |>
    filter(!grepl("nationale|offres", locality),
           !is.na(locality))
}

```
````

might not need to be unit-tested. An assertion, which we will learn about in the
next section, is likely better suited. However, as functions become more
complex, unit tests are highly recommended. This is because it can become very
difficult to make sure that changing some part of the function somewhere does
not affect some other part. This is where writing several unit tests can be
useful. As long as all unit tests keep succeeding (or passing) you are somewhat
sure that what you’re doing is not brekaing stuff. And unit tests are especially
useful when collaborating using trunk-based development! As the project leader,
you could for example refuse to merge changes that break unit tests (the first
project I worked on that used unit tests was organized like this, so we all better
made sure that we wrote sensible code).

Before continuing, let’s rewrite the test we have already. While it is fully
working, we didn’t really write it in the canonical form. Inside
`dev/save_data.Rmd`, change the code of the test to the following:

````{verbatim}
```{r tests-clean_flat_data}
# We now need to check if we have them all in the data. The test needs to be self-contained, hence
# why we need to redefine the required variables:

former_communes <- get_former_communes()

current_communes <- get_current_communes()

communes <- get_test_communes(former_communes, current_communes)

raw_data <- get_raw_data(url = "https://github.com/b-rodrigues/rap4all/raw/master/datasets/vente-maison-2010-2021.xlsx")

flat_data <- clean_raw_data(raw_data)

test_that("Check if all communes are accounted for", {

  expect_true(
    all(communes %in% unique(flat_data$locality))
  )

})
```
````

The only difference is that instead of calling `testthat::expect_true()` directly,
we have wrapped this call inside `testthat::test_that()`. This why, we can add
a description to the test. This is useful if the test fails.

Save `dev/save_data.Rmd` and go back to `0-dev_history.Rmd` to inflate
`save_data.Rmd` again. Everything should work without problems.

If the test fails, you get an informative message. To illustrate, I’ve added a typo
in the test and inflated `save_data.Rmd`. Because tests always run when files
get inflated, this test failed and here is what happened:

```
══ Failed tests ════════════════════════════════════════════════════════════════
── Error ('test-get_raw_data.R:18'): Check if all communes are accounted for ───
Error in `communs %in% unique(flat_data$locality)`: object 'communs' not found
Backtrace:
    ▆
 1. ├─testthat::expect_true(all(communs %in% unique(flat_data$locality))) at test-get_raw_data.R:18:2
 2. │ └─testthat::quasi_label(enquo(object), label, arg = "object")
 3. │   └─rlang::eval_bare(expr, quo_get_env(quo))
 4. └─communs %in% unique(flat_data$locality)

[ FAIL 1 | WARN 2 | SKIP 0 | PASS 0 ]
Error: Test failures
Execution halted
```

The file `test-get_raw_data.R` contains our test, generated by inflating
`save_data.Rmd`. You can find it under the `tests/testthat/` folder of your
inflated package. You can also see the description that we’ve added, which
helps us find the test that failed. In cases like this, you should go back
to the function that makes the test fail and correct it, until the test passes.
You should also make sure that everything is alright with the test itself. If 
there really is a typo in the test, you should of course correct the test 
(in `dev/save_data.Rmd`, not in `tests/testthat/`)!

Now, let’s add a unit test to a complex function, `get_laspeyeres()`. This
function seems to me like a good candidate for testing, as it is not that
trivial.

Let’s try with something simple. `get_laspeyeres()` expects either
`commune_level_data` or `country_level_data`. What happens if we provide another 
dataset? Very likely an error. So let’s test for this. Go back to the
`save_data.Rmd` file and add the following, under the function definition of 
`get_laspeyeres()`:


````{verbatim}
```{r tests-get_laspeyeres}

test_that("Wrong data", {

  expect_error(
    get_laspeyeres(mtcars)
  )

})

```
````

Since we expect an error, we used `expect_error()`, which succeeds if the code
fails! If you’re confused, no worries, we’ve all been there. But let’s think
about it: what would you want to happen if you provided a wrong data set?
Surely, you’d like for the function to scream an error at you, and not somehow
do something and return *something*. So testing that functions fail when they
should is actually quite important as well. Let’s add another, similar, test:

````{verbatim}
```{r tests-get_laspeyeres}

test_that("Wrong data", {

  expect_error(
    get_laspeyeres(mtcars)
  )

})

test_that("Empty data", {

  expect_error(
    get_laspeyeres(subset(mtcars, am == 2))
  )

})

```
````

This second test checks what happens if we provide an empty dataset. This should
not happen, but hey, it’s always a good idea to see what could happen. Here we also
expect an error, se we use `expect_error()` as well. Inflating `save_data.Rmd` 
runs the tests again, all of them successfully.

Now, I know what you’re thinking. Probably something along the lines of 
*"Bruno, you told me that making my projects reproducible and reliable and robust
would not take much more time than what I was already doing before. This certainly
doesn’t feel like it!"* to wish I answer that your feelings on the issue are wrong.
It may not feel like it, but doing this does two things:

- It ultimately saves you time. You typed the test once, and can now rerun it automatically every time you inflate the `.Rmd` files. You don’t need to remember to test the code, and don’t need to remember how to test the code.
- This saves you a lot of headaches. You don’t have to live in fear that you might forget to test the code, or forget how to test the code. You wrote the tests down, and now you’re free to concentrate on adding features or using the existing code knowing that you can trust its outputs.

Trust the process.

Let’s go back to the two tests from before: `get_laspeyeres()` fails, as expected,
when we provide a random dataset to it. But it would be interesting to know why
it fails. Simply run `get_laspeyeres(mtcars)` in the console. This is what we get
back:

```{r, eval = F}
Error in `mutate()`:
           ! Problem while computing `p0 = ifelse(year == "2010",
  average_price_nominal_euros, NA)`.
Caused by error in `ifelse()`:
! object 'year' not found
Run `rlang::last_error()` to see where the error occurred.
```

So the functions fails, but actually, for the wrong reason. It fails because a
column `year` cannot be found in the data. But what if there was a column year?
The code would continue, but then likely fail for something else. It would be
much safer to make it fail as soon as it detects that the provided data sets are
not one of `commune_level_data` and `country_level_data`. But for this, we need
assertive programming, which we will discuss in the next section. Remember, unit
testing tests should run during development time, and assertive testing is for
run-time. In the next section we will be changing the function to fail when the
right datasets are not provided, but our unit test will not need to change; the
function still fails, but it’ll be for the right reasons.

This is another advantage of writing unit tests: it forces you to think about
what you’re doing. It very often improves your code quite a lot, and not just
from a pure algorithmic perspective, but also from a user experience
perspective. Writing these tests made use think about the failure of our
function when we provide a random dataset, and made us realise that it would be
much better for users if the returned error message is something such as "Wrong
dataset, please provide either `commune_level_data` or `country_level_data`".

Let’s continue with testing `get_laspeyeres()`. It would be nice to see if the
function actually does what it’s supposed to do correctly. For this, we need to
start from an input, and then create the expected output. It doesn’t matter how
you create this output, what matters is that you make absolutely sure that it is
correct, and then, never touch it ever again. Let’s call this output the
"truth". Then, you use `get_laspeyeres()` or any other function that you want to
test, and give it this input. You then compare the "truth" to this output. If 
everything matches, congratulations, your function produces the right output.

So let’s start. Remember that unit tests should be self-contained, so I’m going
to create the input dataset and the expected data set (what I called the "truth")
in the test itself. This is the code I’m going to use to create the mock, input
dataset:

```{r, eval = F}
input_df <- expand.grid(
  list("year" = c(2010, 2011),
       "locality" = c("Bascharage", "Luxembourg"))
)

input_df$n_offers <- c(123, 101, 1230, 1010)
input_df$average_price_nominal_euros <- c(234, 345, 560, 670)
input_df$average_price_m2_nominal_euros <- c(23, 34, 56, 67)
```

This creates a data frame with two years, two communes and some mock prices. 
Now, I need to create the output. I start from the input, and add the columns
that `get_laspeyeres()` computes "by hand". Remember, you need to make sure
that these results are correct!

```{r, eval = F}
  expected_df <- input_df

  # p0 should be always equal to the value in the first year
  expected_df$p0 <- c(234, 234, 560, 560)
  expected_df$p0_m2 <- c(23, 23, 56, 56)

  # pl should be equal to the price divided by p0
  expected_df$pl <- expected_df$average_price_nominal_euros/expected_df$p0 * 100
  expected_df$pl_m2 <- expected_df$average_price_m2_nominal_euros/expected_df$p0_m2 * 100
```

If you look at each line, you see that this is basically what `get_laspeyeres()` does. 

````{verbatim}
```{r, eval = F}
test_that("get_laspeyeres() produces correct results", {

  input_df <- expand.grid(
    list("year" = c(2010, 2011),
         "locality" = c("Bascharage", "Luxembourg"))
  )

  input_df$n_offers <- c(123, 101, 1230, 1010)
  input_df$average_price_nominal_euros <- c(234, 345, 560, 670)
  input_df$average_price_m2_nominal_euros <- c(23, 34, 56, 67)

  expected_df <- input_df

  # p0 should be always equal to the value in the first year
  expected_df$p0 <- c(234, 234, 560, 560)
  expected_df$p0_m2 <- c(23, 23, 56, 56)

  # pl should be equal to the price divided by p0
  expected_df$pl <- expected_df$average_price_nominal_euros/expected_df$p0 * 100
  expected_df$pl_m2 <- expected_df$average_price_m2_nominal_euros/expected_df$p0_m2 * 100

  expect_equivalent(
    expected_df, get_laspeyeres(input_df)
  )

})
```
````



## Assertive programming

The analysis is still in Quarto, so how could the readers of this book test their code?
Copying here what Miles wrote on the subject:

*‘Assertive programming’ is a topic that might be missing from the book. I think
of it as a kind of dual of unit testing. Unit testing is for more generally
applicable packaged code. But when you have functions in your analysis pipeline
that operate on a very specific kind of input data, unit testing becomes kind of
nonsensical because you’re left to dream up endless variations of your input
dataset that may never occur. It’s a bit easier to flip the effort to validating
the assumptions you have about your input and output data, which you can do in
the pipeline functions themselves rather than separate unit testing ones. This
is nice because it ensures the validation is performed in the pipeline run, and
so is backed by the same reproducibility guarantees.*

## Test-driven development

## Code coverage

