---
title: "Templating with child documents"
output: html_document
date: "2023-01-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
```

## A function that creates ggplots

```{r}
create_plot <- function(dataset, ...){
  ggplot(dataset) +
    geom_point(...)

}
```

The function above takes a dataset and an aesthetic made using `ggplot2::aes()` to
create a plot:

```{r}
create_plot(mtcars, aes(y = mpg, x = hp))
```

Let’s suppose that we want to generate a document that would look like this:

- first a section title, with the aesthetic used;
- then a plot

So it would look like this:

## Plot with aesthetic: aes(y = mpg, x = hp)

```{r}
create_plot(mtcars, aes(y = mpg, x = hp))
```

We don’t want to create these sections for every variable by hand.

Instead, we can make use of a child document that gets knitted separately
and then embedded in the parent document. This is the child document:

```{r, results = "asis"}

res <- lapply(list(aes(y = mpg, x = hp),
                aes(y = mpg, x = hp, size = am)),
              function(dataset, x){

  knitr::knit_child(text = c(

    '\n',
    '## Chosen aesthetic: `r dQuote(x)`',
    '\n',
    '```{r, echo = F}',
    'print(create_plot(dataset, x))',
    '```'

     ),
     envir = environment(),
     quiet = TRUE)

}, dataset = mtcars)


cat(unlist(res), sep = "\n")
```

The child document is the `text` argument to the `knit_child()` function.
`text` is literal R Markdown code: we define a level 2 header, and then an 
R chunk. This child document gets knitted, so we need to specify the 
environment that should get used. This means that the child document will
get knitted in the same environment as the parent document (our current 
global environment). This way, every package that get loaded and every function
or variable that got defined in the parent document will also be available to
the child document.

We then use `lapply()` to loop over two aesthetics with an anonymous function
that encapsulates the child document. So we get two child documents that get
knitted, one per aesthetic. This gets saved into variable `res`. This is thus
a list of knitted Markdown.

Finally, we need unlist `res` to actually merge the Markdown code from the 
child documents into the parent document.
